---
title: "An Introduction to `adelie`"
author:
    - James Yang and Trevor Hastie
date: "`r format(Sys.time(), '%B %d, %Y')`"
bibliography: assets/grpnet_refs.bib
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{An Introduction to `adelie`}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "##",
  fig.width = 7,
  fig.height=6
)
```
# __Introduction to Group Elastic Net__

The R package `adelie` is built from the same C++ code as used in the
corresponding [adelie Python package](https://github.com/JamesYang007/adelie).

In this notebook, we give a brief overview of the group elastic net problem that `adelie` solves.


## __Single-Response Group Elastic Net__


The single-response group elastic net problem is given by
$$
\begin{align*}
    \mathrm{minimize}_{\beta, \beta_0} \quad&
    \ell(\eta) + \lambda \sum\limits_{g=1}^G \omega_g \left(
        \alpha \|\beta_g\|_2 + \frac{1-\alpha}{2} \|\beta_g\|_2^2
    \right)
    \\\text{subject to}\quad&
    \eta = X \beta + \beta_0 \mathbf{1} + \eta^0
\end{align*}
$$
where 
$\beta_0$ is the intercept,
$\beta$ is the coefficient vector,
$X$ is the feature matrix,
$\eta^0$ is a fixed offset vector,
$\lambda \geq 0$ is the regularization parameter,
$G$ is the number of groups,
$\omega_g \geq 0$ is the penalty factor per group,
$\alpha \in [0,1]$ is the elastic net parameter,
and $\beta_g$ are the coefficients for the $g$ th group.
$\ell(\cdot)$ is the loss function defined by the GLM.
As an example, the Gaussian GLM 
(`glm.gaussian()`)
defines the loss function as
$$
\begin{align*}
    \ell(\eta)
    &=
    \sum\limits_{i=1}^n w_i \left(
        -y_i \eta_i + \frac{\eta_i^2}{2}
    \right)
\end{align*}
$$
where
$w \geq 0$ is the observation weight vector,
$y$ is the response vector,
and $\eta$ is the linear prediction vector as in the optimization
problem above. This is equivalent to the weighted sum-of-squared errors
$$
\begin{align*}
    \mbox{RSS}
    &=
    \frac12\sum\limits_{i=1}^n w_i \left(
        y_i -\eta_i \right)^2,
\end{align*}
$$
since the term in $\sum w_iy_i^2$ is a constant for the
optimization.

Specifically for the Gaussian GLM, we employ a specialized optimizer based on coordinate descent
to solve the group elastic net problem.

The Gaussian GLM is written in "exponential family" form, with
$\eta_i$ the natural parameter. Other GLMs such as binomial
(`glm.binomial()`) and Poisson (`glm.poisson()`) have similar
expressions for the negative log-likelihood.
For other general GLMs as well as the Cox model (`glm.cox()`), we use a proximal Newton method, 
which leads to an Iterative Reweighted Least Squares (IRLS) algorithm,
That is, we iteratively perform a quadratic approximation to $\ell(\cdot)$, 
which yields a sequence of Gaussian GLM group elastic net problems
that we solve using our special solver based on coordinate descent.


The Gaussian GLM also admits a different algorithm, which we call the _the covariance method_,
using summary statistics rather than individual-level data.
The covariance method solves the following problem:
$$
\begin{align*}
    \mathrm{minimize}_{\beta} \quad&
    \frac{1}{2} \beta^\top A \beta
    - v^\top \beta
    + 
    \lambda \sum\limits_{g=1}^G \omega_g \left(
        \alpha \|\beta_g\|_2 + \frac{1-\alpha}{2} \|\beta_g\|_2^2
    \right)
\end{align*}
$$
This method would be equivalent to the usual single-response Gaussian group elastic net problem
if $A \equiv X_c^\top W X_c$ and $v \equiv X_c^\top W y_c$
where $X_c$ is column-centered version of $X$ 
and $y_c$ is the centered version of $y-\eta^0$
where the means are computed with weights $W$
(if intercept is to be fit).

This method only works for the Gaussian case since the proximal Newton method
changes the weights $W$ at every IRLS iteration,
so that without access to $X$, it is not possible to compute the new "$A$" and "$v$".


## __Multi-Response Group Elastic Net__


The multi-response group elastic net problem is given by
$$
\begin{align*}
    \mathrm{minimize}_{B,\; \beta_0} \quad&
    \ell(\eta) + \lambda \sum\limits_{g=1}^G \omega_g \left(
        \alpha \|B_g\|_F + \frac{1-\alpha}{2} \|B_g\|_F^2
    \right)
    \\\text{subject to}\quad&
        \eta = X B + \mathbf{1} \beta_0^\top + \eta^0
\end{align*}
$$
The model arises in "multitask" learning --- `glm.multigaussian()` ---
as
well as multinomial (multiclass) logistic regression
--- `glm.multinomial()`. This way, we have possibly different linear
predictions for each of $K$ responses, and hence 
$\eta$ is $n\times K$. The coefficient "matrices" for each group $B_g$ are
penalized using a Frobenius norm. 
Note that if an intercept is included in the model, an intercept is added for each response.


We use an alternative but equivalent representation in the `adelie` software, by
"flattening" the coefficient matrices. Specifically we solve
$$
\begin{align*}
    \mathrm{minimize}_{\beta,\; \beta_0} \quad&
    \ell(\eta) + \lambda \sum\limits_{g=1}^G \omega_g \left(
        \alpha \|\beta_g\|_2 + \frac{1-\alpha}{2} \|\beta_g\|_2^2
    \right)
    \\\text{subject to}\quad&
    \mathrm{vec}(\eta^\top) = (X \otimes I_K) \beta + (\mathbf{1} \otimes I_K) \beta_0 + \mathrm{vec}(\eta^{0\top})
\end{align*}
$$
where $\mathrm{vec}(\cdot)$ is the operator that flattens a column-major matrix into a vector,
and $A \otimes B$ is the Kronecker product operator. Hence $\beta \equiv \mathrm{vec}(B^\top)$.

As indicated above, the multi-response group elastic net problem is technically of the same form
as the single-response group elastic net problem.
In fact, `adelie` reuses the single-response solver for multi-response problems
by modifying the inputs appropriately 
(e.g. using `matrix.kronecker_eye()`) to represent $X \otimes I_K$).
For the MultiGaussian family, we wrap the specialized single-response Gaussian solver
and otherwise for general multi-response GLMs, we wrap the single-response GLM solver.


# __Quickstart__

In this section, we cover the basic usage of `adelie`.

```{r}
library(adelie)
```

## __Gaussian Group Elastic Net__

Before using `adelie`, we assume that the user has a feature matrix `X` and a response vector `y`.
For simplicity, we assume for now that `X` is a dense matrix,
although we will later see that `X` can be substituted with a custom matrix as well.
For demonstration, we will randomly generate our data.

```{r}
n = 100     # number of observations
p = 1000    # number of features
set.seed(5) # seed
X = matrix(rnorm(n*p),n,p)
y = X[,1:10]%*%rnorm(10) + rnorm(n)*sqrt(10) # makes SNR = 1
```

### __Lasso__


The most basic call to `grpnet` simply supplies a `X` matrix and a `glm` object.
For example, `glm.gaussian()` returns a Gaussian `glm` object,
which corresponds to the usual least squares loss function.
By default, `grpnet` will fit lasso by setting each feature as its own group.
`adelie` is written to treat groups of size `1` in a more optimized manner,
so it is a competitive lasso solver that has computation times similar
to those of the `glmnet` package.
Like `glmnet`, `grpnet` will also generate a sequence of 100  regularizations $\lambda$ 
evenly spaced on the log scale, by default if the user does not provide the path.
Since `grpnet` is a path-solver, it will warm-start at the next $\lambda$
using the current solution on the path.
__For this reason, we recommend users to supply a sufficiently fine grid of__ $\lambda$ __or use the default path!__

```{r}
fit = grpnet(X=X, glm=glm.gaussian(y=y))
print(fit)
```

The `print()` methods gives a nice summary of the fit.
Note that the solver finished early in this example.
By default, the solver terminates if the latter reaches $90\%$ as a simple heuristic to avoid overfitting.
This threshold can be controlled via the argument `adev_tol`.

The output of `grpnet` is a an object of class "grpnet", which
contains some simple descriptors of the model, as well as a _state_
object that represents the state of the optimizer.
For most use-cases, the users do not need to inspect the internals of a state object,
but rather can extract the useful information via the methods
`print()`, `plot()`, `predict()` and `coef()`.

```{r}
plot(fit)
```
Here we plot the coefficients profiles as a function of
$-\log(\lambda)$. By default `grpnet` standardizes the features
internally, and these coefficients are on the scale of the
standardized features. One can avoid standardization via `standardize
= FALSE` in the call to `grpnet()`.

We can make predictions at new values for the features --- here we use
a subset of the original:

```{r}
pred = predict(fit, newx = X[1:5,],lambda = c(1.5, 1))
pred
```
Finally, we would like to choose a good value for $\lambda$, and for
that we use cross-validation. We include a `progress_bar` argument,
which can be helful for big problems. The `plot` method for a
`cv.grpnet` object displays the average cross-validated deviance of the model,
with approximate standard error bars for the average. 

```{r}
fitcv = cv.grpnet(X,glm.gaussian(y),progress_bar = TRUE)
plot(fitcv)
```
Two vertical
dashed lines are included: one at the value of $\lambda$ corresponding
to the minimum value, and another at a larger (more conservative)
value of $\lambda$, such that the mean error is within one standard
error of the minimum.

One can print the fitted `cv.glmnet` object:
```{r}
fitcv
```
One can also predict directly from it:
```{r}
pred = predict(fitcv, newx = X)
pred = predict(fitcv, newx = X, lambda = "lambda.min")
```
In the first case the prediction is at the default value of $\lambda$,
which is `lambda = "lambda.1se"`. Alternatively, any numeric values of
$\lambda$ can be supplied.

### __Group Lasso__


To fit group lasso, the user simply needs to supply a `groups` argument that
defines the starting column index of $X$ for each group.
For example, if there are `4` features with two (contiguous) groups of sizes `3` and `1`, respectively,
then `groups = c(1, 4)`.
For demonstration, we take the same data as before and group every `10` features.
We then fit group lasso using the same function.

```{r}
fitg = grpnet(
    X=X,
    glm=glm.gaussian(y=y),
    groups=seq(from = 1, to = p, by=10),
    )
print(fitg)
plot(fitg)
```
Notice in the printout the active set (df) increases in steps of 10,
as expected. Also the coefficient plot colors all coefficients for a
group with the sample color.


As before, we can use cross-validation to select the tuning parameter.
```{r}
fitgcv = cv.grpnet(
    X,
    glm.gaussian(y),
    groups=seq(from = 1, to = p, by=10),
    progress_bar = TRUE)
plot(fitgcv)
```


## __GLM Group Elastic Net__

In the previous section, we covered how to solve penalized least squares regression.
Nearly all of the content remains the same for fitting penalized GLM regression.
The only difference is in the choice of the `glm` object.
For brevity, we only discuss logistic regression as our non-trivial GLM example,
however the following discussion applies for any GLM.

Let's modify our example and generate a binomial response.

```{r}
eta = X[,1:5]%*%rnorm(5)/sqrt(5)
mu = 1 / (1 + exp(-eta))
y = rbinom(n, size = 1, prob = mu)
```

To solve the group elastic net problem using the logistic loss,
we simply provide the binomial GLM object.
For simplicity, we fit the lasso problem below.

```{r}
fitb = grpnet(X, glm.binomial(y))
plot(fitb)
```

Cross-validation works as before:
```{r}
fitbcv = cv.grpnet(X,glm.binomial(y),progress_bar = TRUE)
plot(fitbcv)
```
We can make predictions from the fitted objects as before. For GLMs,
the predictions are for the link ($\eta$).

We can specify coefficient groups just as before.

## __Multi-Response GLM Elastic Net__

`grpnet` is also able to fit multi-response GLM elastic nets.
Currently, we only support MultiGaussian (least squares) and Multinomial GLMs.
 
The following code shows an example of fitting a multinomial
regression problem. We will use the `glm.multinomial()` family, which
expects a matrix response $Y$ with $K$ columns (the number of
classes). Each row of $Y$ is a proportion vector which sums to 1.
In the example below, $Y$ is an indicator matrix, with a single 1 in
each row, the rest being zeros.

```{r}
n = 300     # number of observations
p = 100    # number of features
K = 4       # number of classes
set.seed(7)
X = matrix(rnorm(n*p),n,p)
eta = X[, 1:5] %*% matrix(rnorm(K*5),5,K)/sqrt(5)
probs = exp(eta)
probs = probs/rowSums(probs)
Y = t(apply(probs,1,function(prob)rmultinom(1, 1, prob)))
Y[1:5,]
```

The fitting proceeds as before. We will directly fit a group lasso,
with groups of 5 chosen similarly to before.

```{r fitm}
grps = seq(from=1, to=p, by = 5)
fitm = grpnet(X, glm.multinomial(Y), groups=grps)
```

```{r  out.lines = 10}
print(fitm)
```

```{r plotfitm}
plot(fitm)
```
The coefficient for each feature is a vector(one per class), so rather
than show multiple coefficients, the plot shows the progress of the
2norm as a function of $\lambda$. Once again, because of the grouping
(into groups of 5 here), the groups are colored the same. In this
case, the first group has all the action, so appears much earlier than
the rest.

```{r}
fitmcv = cv.grpnet(X,glm.multinomial(Y),groups=grps)
plot(fitmcv)
```

Another important difference from the single-response case is that 
the user must be aware of the shape of the returned coefficients, as
returned by `coef(fitm)` or `coef(fitmcv)`. 
```{r}
names(coef(fitm))
```
We see that `coef()` returns a list.
For multi-response GLMs, the `intercepts` component is
included by default for each response,
and hence is a $L \times K$ matrix, where $L$ is the number of
$\lambda$s in the path. The `betas` component will be a $L \times pK)$
sparse matrix where every successive $K$ columns correspond to the
coefficients associated with a feature and across all responses.

Fortunately the `predict()` method understands this structure, and
behaves as intended.

```{r}
predict(fitmcv,newx = X[1:3,])
```
Here by default the prediction is made at `lambda = "lambda.1se"`.

## __Cox Models__

Our final example will be for censored survival data, and we will fit
Cox proportional hazards model.

We create an example with a $500 \times 100$ _sparse_ $X$ matrix.
In this example we have right censoring. So the "subject" exits at
times in `y`, and the binary `status` is 1 of the exit is a "death",
else 0 if censored.

```{r}
set.seed(9)
n <- 500
p <- 100
X  <- matrix(rnorm(n*p), n, p)
X[sample.int(n * p, size = 0.5 * n * p)] <- 0
X_sparse <- Matrix::Matrix(X, sparse = TRUE)

nzc <- p / 4
beta <- rnorm(nzc)
fx <- X[, seq(nzc)] %*% beta / 3
hx <- exp(fx)
y <- rexp(n,hx)
status <- rbinom(n = n, prob = 0.3, size = 1)
```

We now fit an elastic-net model using the `glm.cox()` family,
with every set of 5 coefficients in a group.

```{r}
groups = seq(from = 1, to = p, by = 5)
fitcv <- cv.grpnet(X_sparse,
                   glm.cox(stop = y, status = status),
                   alpha = 0.5,
                   groups = groups)
par(mfrow = c(1,2))
plot(fitcv)
plot(fitcv$grpnet.fit)
```


## References
